---
title: "Empirical Data: Detect Epidemics"
author: "Anna Maria Langmüller"
date: "2024-07"
output: html_document
---

# Setup
working directory, packages, functions, global variables. 

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/anna/Dropbox (Personal)/DD/res/20240726/")
library(plyr)
library(ggplot2)
library(tidyverse)
library(cowplot)
library(npreg)
```

```{r functions, include=FALSE}
#calc_empirical_imax: function to calculate max incidence from empirical data 
#x ... row from my_epidemics data frame 
#y ... empirical dengue incidence 
calc_empirical_imax <- function(x, y) {
  y_s <- subset(y, ADM2_PCODE == x[["ADM2_PCODE"]]) #subset to municipality
  
  id_epidemic <- which(y_s$calendar_start_date >= x[["xmin"]] & #subset to epidemic
                         y_s$calendar_start_date <= x[["xmax"]])
  y_s <- y_s[id_epidemic,]
  imax <- max(y_s$dengue_freq)
  return(imax)
}

plot_epidemics <- function(my_epis, my_municipality, my_counts) {
  #plot_epidemics: plot epidemics in specific municipality
    #my_epis: detected epidemics
    #my_municipality: municipality ID
    #my_counts: dengue incidence data 
  
  #filter epidemics for specific municipality:
  my_epi <- my_epis[which(my_epis$ADM2_PCODE == my_municipality),] 
  
  #filter incidence data for specific municipality:
  my_counts <- my_counts[which(my_counts$ADM2_PCODE == my_municipality),]
  
  ss.mod <- ss(x = my_counts$calendar_start_date, #fit spline
               y = my_counts$dengue_freq,
               lambda = LAMBDA)
  
  ss.fit <- data.frame(calendar_start_date = my_counts$calendar_start_date,
                       dengue_freq = fitted(ss.mod)) #predicted dengue_freq
  
  g <- ggplot() + theme_minimal()
  g <- g + geom_bar(data = my_counts,
                    aes(x = calendar_start_date,
                        y = dengue_freq),
                    stat = "identity", fill = "white", color = "grey")
  
  g <- g + geom_hline(yintercept = my_epi[["thres"]], col = "#00cdff")
  g <- g + geom_hline(yintercept = THRES, col = "#00cdff")
  g <- g + geom_rect(data = my_epi, aes(xmin = xmin, xmax = xmax),
                     ymin = -Inf, ymax = Inf, fill = "#00cdff",
                     alpha = .3)
  g <- g + geom_line(data = ss.fit, aes(x = calendar_start_date, y = dengue_freq),
                     col = "#de5285")
  g <- g + xlab("Date")
  g <- g + ylab("Adj. prop. dengue incidences per week")
  g <- g + ggtitle(my_municipality)
  return(g)
}

filter_min_count_epidemics <- function(df, min_count = 3) {
  df_filtered <- df %>%
    group_by(ADM2_PCODE) %>%
    filter(n() >= min_count) %>%
    ungroup()
  return(df_filtered)
}

# Function to split the dataframe into training and testing sets
split_dataframe <- function(df, train_proportion = 2/3) {
  df_train <- data.frame() #training data
  df_test <- data.frame() # test data 
  
  groups <- df %>% group_by(ADM2_PCODE) %>% group_split()
  
  for (group in groups) {
    num_outbreaks <- nrow(group)
    indices <- sample(1:num_outbreaks) # Randomly shuffle indices
    split_index <- floor(train_proportion * num_outbreaks) # Calculate split index
    
    df_train <- bind_rows(df_train, group[indices[1:split_index], ])
    df_test <- bind_rows(df_test, group[indices[(split_index + 1):num_outbreaks], ])
  }
  return(list(df_train, df_test))
}

```

```{r global_variables, include=FALSE}
PATH_OpenDengue<- "../../data/20240612/Highest temporal resolution data_COLOMBIA_20021229_20191231.csv" #dengue incidence
PATH_Demography <- "../../data/20240220/doi_10_5061_dryad_83nj1__v20190404/spatial_aggregate_non_timeseries/municip_aggregate_non_ts.csv" #human demography
PATH_Aedes <- "../../data/20240220/doi_10_5061_dryad_83nj1__v20190404/weighted_spatial_aggregates_municip/municip_Ae_aegypti_weeks_weighted.csv" #aedes scores 
PATH_Link <- "../20240725/DD-AML-res-IDLinks.txt" #link data 

THRES <- 0.001 #max incidence threshold 
N_THRES <- 30000 #minimum number of inhabitants/municipality
CORR_COEF <- 25 #correction factor for dengue cases 
PROP_PLOT <- 0.05 #proportion of epidemics that should be displayed (--> .html generation)
LAMBDA <- 1e-10 #lambda parameter for spline regression
```

# Analysis Purpose

The purpose of this analysis is to detect epidemic dengue outbreaks for Colombia between 2007 and 2019. The dengue incidence data was obtained from the OpenDengue database.

```{r prep_data, include=FALSE}

dengue <- read.csv(PATH_OpenDengue) #read data 
nonInformative <- apply(dengue, 2, function(x) length(unique(x))) #determine number of distinct values per column
for (i in names(nonInformative)[which(nonInformative == 1)]) #remove non-informative columns (i.e., only 1 value)
  dengue[[i]] <- NULL
rm(nonInformative, i)
dengue$calendar_start_date <- as.Date(dengue$calendar_start_date)
dengue$calendar_end_date <- as.Date(dengue$calendar_end_date)

pop <- read.csv(PATH_Demography) #demography of municipals
aedes <- read.csv(PATH_Aedes) #mosquito abundance 

link <- read.table(PATH_Link, header = T) #link Open Dengue vs. Siraj
link$ID_ESPACIA <- as.integer(gsub("CO","", link$ADM2_PCODE)) #reformat ID for Siraj data

#subset to linked municipalities
dengue <- merge(dengue, subset(link, select = c("ID_ESPACIA", "full_name","ADM2_PCODE")),
                by = "full_name")
pop <- merge(pop, subset(link, select = c("ID_ESPACIA", "ADM2_PCODE")),
             by = "ID_ESPACIA")  
aedes <- merge(aedes, subset(link, select = c("ID_ESPACIA", "ADM2_PCODE")),
               by = "ID_ESPACIA")
```

# Data Overview

If we only consider municipalities for which I was able to link data between the Open Dengue data base and the Siraj et al. (2018) paper (see 2024-07-25), there are 154,060 data entries in the Open Dengue database between 2007 and 2019. The highest available temporary resolution is weeks, resulting in 678 reporting periods in 1,009 different municipalities. Reporting periods are only included into the database if the number of reported dengue cases is at least 1 for the respective week and municipal. On average, there are 153 reporting periods per municipal. The reported incidence ranges between 1 and 839 cases per week and municipal.

```{r data_overview, include = FALSE}
if(any(dengue$T_res!= "Week")) #remove any coarse temporal entries 
  dengue <- subset(dengue, T_res == "Week")

nrow(dengue) #no. data entries 
length(unique(dengue$calendar_end_date)) #number of different end dates 
length(unique(dengue$calendar_start_date)) #number of different start dates
length(unique(dengue$ADM2_PCODE)) #number of different municipal

table(as.numeric(as.Date(dengue$calendar_end_date) - as.Date(dengue$calendar_start_date))) #reporting periods 
dataM <- ddply(dengue , .(ADM2_PCODE), summarize, n = length(unique(calendar_start_date))) #no. entries / municipal 
summary(dataM$n)
dataM <- ddply(dengue, .(ADM2_PCODE), summarize, maxInc = max(dengue_total)) #maximum incidence / municipal 
summary(dataM$maxInc)
rm(dataM)
```

```{r fill_sparse_date, include=FALSE}
all_municipalities <- subset(dengue, select = c("full_name", #unique municipality info
                                                "adm_1_name",
                                                "adm_2_name",
                                                "FAO_GAUL_code",
                                                "RNE_iso_code",
                                                "ID_ESPACIA",
                                                "ADM2_PCODE"))
all_municipalities <- all_municipalities[!duplicated(all_municipalities),]

all_weeks <- subset(dengue, select = c("calendar_start_date", #unique time info 
                                       "calendar_end_date",
                                       "Year"))
all_weeks <- all_weeks[!duplicated(all_weeks),]

template_df <- expand.grid(ADM2_PCODE = all_municipalities$ADM2_PCODE, calendar_start_date = all_weeks$calendar_start_date) #create template df 
template_df <- merge(all_municipalities, template_df, by = "ADM2_PCODE")
template_df <- merge(template_df, all_weeks, by = "calendar_start_date")

dengue_df <- merge(template_df, dengue, #fill in incidence data 
                     by = c("full_name",
                            "adm_1_name",
                            "adm_2_name",
                            "FAO_GAUL_code",
                            "RNE_iso_code",
                            "ID_ESPACIA",
                            "ADM2_PCODE", 
                            "calendar_start_date",
                            "calendar_end_date",
                            "Year"), all.x = TRUE)

dengue_df$dengue_total[is.na(dengue_df$dengue_total)] <- 0 #fill missing incidences
dengue_df <- dengue_df %>% arrange(ADM2_PCODE, calendar_start_date) #sort

table(table(dengue_df$ADM2_PCODE)) #sanity check
table(table(dengue_df$calendar_start_date)) #sanity check
rm(all_municipalities, all_weeks, template_df)
gc()
```

# Detecting Epidemics

We will focus on municipals that (a) roughly fit our agent-based simulations in terms of population size (i.e., at least 30,000 individuals), and (b) have an adjusted maximum incidence of at least 0.1 % in any of the reported periods. For the adjusted dengue incidences, we correct the reported dengue incidences by a factor of 25. We acknowledge this is a rather crude simplification.

211 municipals pass these filtering criteria. The bar plots below display adjusted dengue incidence proportions per week between 2007 and 2019. To detect epidemics, we fitted a smoothing spline using the ss() function from the npreg R-package with a lambda of 1e-10. An epidemic outbreak was defined as a time period of at least 4 weeks in which the spline function exceeded the median adjusted dengue incidence count. 

Using this approach, we discovered 1289 potential epidemic outbreaks. 1211 out these 1289 outbreaks had an adjusted empirical maximal incidence of at least 0.1 % and were considered for further analysis. The plots below show some examples of this approach: Epidemic outbreaks are highlighted as light-blue rectangles. 

```{r detect_epidemics_prep, echo=FALSE}

dengue_df <- merge(dengue_df, subset(pop, select = c("ID_ESPACIA", "Wpop2015"))
                   , by = "ID_ESPACIA", all.x = T)

#expansion factor and calculate proportions 
dengue_df$dengue_freq <- dengue_df$dengue_total * CORR_COEF / dengue_df$Wpop2015 

candidate_id <- which(dengue_df$dengue_freq >= THRES 
                      & dengue_df$Wpop2015 >= N_THRES)

candidate_id <- unique(dengue_df$ADM2_PCODE[candidate_id]) #municipality candidates for epidemics 

my_epidemics <- data.frame(xmin = as.Date(character()), #prep my_epidemics df
                           xmax = as.Date(character()),
                           thres = numeric(),
                           ADM2_PCODE = character(),
                           epidemic_id = character())
```


```{r detect_epidemics, include=FALSE}
for (i in candidate_id) { #for each candidate municipality
  dengue_iter <- subset(dengue_df, ADM2_PCODE == i) #subset data
  #calculate median incidence:
  thres_iter <- quantile(dengue_iter$dengue_freq[dengue_iter$dengue_freq > 0], 0.5) 
  ss.mod <- ss(x = dengue_iter$calendar_start_date, #fit spline
               y = dengue_iter$dengue_freq,
               lambda = LAMBDA)
  
  ss.fit <- data.frame(calendar_start_date = dengue_iter$calendar_start_date,
                       dengue_freq = fitted(ss.mod)) #predicted dengue_freq
  ss.fit$epidemic <- ifelse(ss.fit$dengue_freq > thres_iter, T, F) #epidemic = T/F
  
  ss.fit$epidemic_id <- 0 #ID epidemics 
  epidemic_id <- 1 #epidemic id counter
  row_counter <- 2 #row counter (while loop)
  
  #look at 1st data entry: 
  ss.fit$epidemic_id[1] <- ifelse(ss.fit$dengue_freq[1] > thres_iter, epidemic_id, 0)
  
  while (row_counter <= nrow(ss.fit)) { #assign epidemic IDs
    if (ss.fit$epidemic[row_counter] == TRUE) {
      if (ss.fit$epidemic[(row_counter - 1)] == FALSE & any(ss.fit$epidemic_id > 0)) {
        epidemic_id <- epidemic_id + 1
      }
    ss.fit$epidemic_id[row_counter] <- epidemic_id
    }
  row_counter = row_counter + 1
  }
  
  if (any(ss.fit$epidemic_id > 0)) { #if epidemics detected
    iter_epidemics <- ss.fit[which(ss.fit$epidemic_id > 0),] #epidemic ranges 
    iter_epidemics <- ddply(iter_epidemics, .(epidemic_id), summarize, 
                            xmin = min(calendar_start_date),
                            xmax = max(calendar_start_date))
    iter_epidemics <- subset(iter_epidemics, xmax - xmin > 27) #at least 4 weeks duration 
    
    if (nrow(iter_epidemics) > 0 ) {
      iter_epidemics$ADM2_PCODE <- i #municipality identifier
      iter_epidemics$epidemic_id <- paste(
        iter_epidemics$ADM2_PCODE, iter_epidemics$epidemic_id, sep = ".")
      iter_epidemics$thres <- thres_iter
      my_epidemics <- rbind(my_epidemics, 
                            subset(iter_epidemics, #store identified epidemics
                                             select = c("xmin", "xmax", 
                                                        "ADM2_PCODE", 
                                                        "epidemic_id", "thres")))
      }
    }
  }
      
print(nrow(my_epidemics)) #raw number of epidemics 

my_epidemics$duration <- my_epidemics$xmax - my_epidemics$xmin #calc. duration 

imax <- sapply(1:nrow(my_epidemics), function(k) #calc. max incidence 
  calc_empirical_imax(x = my_epidemics[k,], y = dengue_df))
my_epidemics$imax <- imax
rm(imax)
my_epidemics <- my_epidemics[-which(my_epidemics$imax < THRES),]  
print(nrow(my_epidemics)) #filtered number of epidemics 

rm(iter_epidemics, ss.fit, ss.mod, epidemic_id, i, dengue_iter, row_counter, thres_iter)
gc()
```


```{r plot_epidemics, echo=FALSE}

sample_candidate_ids <- sample(x = candidate_id, size = 10)

for (id in sample_candidate_ids)
  print(plot_epidemics(my_epis = my_epidemics,
                 my_municipality = id,
                 my_counts = dengue_df))

rm(sample_candidate_ids, id)
```

# Characterizing Epidemics

On average, each municipality had 6.34 outbreaks. The average duration per outbreak was 195 days, with an average maximum incidence of 0.6 %. 
```{r describe_epidemics, include=FALSE}
mean(table(my_epidemics$ADM2_PCODE))
mean(my_epidemics$duration)
mean(my_epidemics$imax)
```
 
 
```{r store_epidemics, include=FALSE}
to_store <- data.frame( #for Fit & Predict approach
  "ADM2_PCODE" = my_epidemics$ADM2_PCODE,
  "epidemic_id" = my_epidemics$epidemic_id,
  "t" = yday(as.Date(my_epidemics$xmin)),
  "imax" = my_epidemics$imax,
  "duration" = as.numeric(my_epidemics$duration)
)

#write.table(my_epidemics, file = "OpenDengue-epidemics.txt",
            row.names = F, col.names = T, quote = F, sep = "\t")

#write.table(to_store, file = "OpenDengue-epidemics-reformat.txt",
            row.names = F, col.names = T, quote = F, sep = "\t")
```


