---
title: "Empirical Data Prep: OpenDengue & Siraj et al. (2018)"
author: "Anna Maria Langmüller"
date: "2024-07"
output: html_document
---

# Setup
working directory, packages, functions, global variables. 

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
WD <- "/Users/anna/Dropbox (Personal)/DD/res/20240725"
setwd(WD)

library(sf) #to read shape files
library(stringi) #to standardize strings
library(stringdist) #to calculate distance between strings 
library(plyr) #for data frame manipulations
library(ggplot2) #for plotting 
```

```{r functions, include=FALSE}

plot_gaul <- function(gaul_id, od_shape, ocha_shape, dengue_data){
  #gaul_id = GAUL CODE identifier
  #od_shape = Open Dengue Shape file, municipality level
  #ocha_shape = OCHA Shape file, municipality level
  #dengue_data = Open Dengue Incidence data 
  
  #standardize municipality names for comparison 
  ocha_shape$adm_1_name <- toupper(
    stri_trans_general(ocha_shape$ADM1_ES, "latin-ascii"))
  ocha_shape$adm_2_name <- toupper(
    stri_trans_general(ocha_shape$ADM2_ES, "latin-ascii"))
  
  od_munic <- od_shape[which(od_shape$GAUL_CODE == gaul_id), ] #subset OpenDengue shape file
  
  #look up municipalities from dengue incidence data 
  dengue_munic <- dengue_data[which(dengue_data$FAO_GAUL_code == gaul_id),]
  dengue_munic <- subset(dengue_munic, select = c("adm_1_name", "adm_2_name"))
  dengue_munic <- dengue_munic[!duplicated(dengue_munic),]
  
  #subset OCHA data 
  ocha_munic <- ocha_shape[which(ocha_shape$adm_2_name %in% dengue_munic$adm_2_name),]
  ocha_munic <- ocha_munic[which(ocha_munic$adm_1_name %in% dengue_munic$adm_1_name),]
  
  # Convert to data frames
  od_munic <- st_as_sf(od_munic)
  ocha_munic <- st_as_sf(ocha_munic)
  
  g <- ggplot() + #plot
  geom_sf(data = ocha_munic, color = "darkblue", fill = NA, linewidth = 1) +
  geom_sf(data = od_munic, color = "deeppink2", fill = NA, linewidth = 1)
  g <- g + theme_minimal()
  g <- g + theme(text = element_text(size = 14))
  g <- g + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
  g <- g + ggtitle(paste("GAUL ID: ", gaul_id))
  return(g)
}

link_data <- function(x, xshape, y) {
  # x ... row from dengue_link data frame
  # xshape ... shape file from OpenDengue project
  # y ... shape file from OCHA
  
  #calculate distance between municipality names 
  y$adm2Score <- sapply(c(1:nrow(y)), function(k)
  stringdist(x[["adm2_std"]], y[["adm2_std"]][k]))
  
  #subset to entries with least difference
  y <- y[which(y$adm2Score == min(y$adm2Score)), ]  
  
  #calculate distance between department names
  y$adm1Score <- sapply(c(1:nrow(y)), function(k)
    stringdist(x[["adm1_std"]], y[["adm1_std"]][k]))
  
  #subset to entries with least difference 
  if (nrow(y) > 1)
    y <- y[which(y$adm1Score == min(y$adm1Score)),]

  y$dengueID <- x$ID #add dengue ID
  
  return(y) #return candidates from OCHA shape file
}

plot_gaul_byIDs <- function(od_id, ocha_id, od_shape, ocha_shape) {
  #od_id ... Open Dengue ID (GAUL CODE)
  #ocha_id ... OCHA municipality id 
  #od_shape ... Shape file from Open Dengue
  #ocha_shape ... Shape file from OCHA 
  
  #subset to single municipality
  od_munic <- od_shape[which(od_shape$GAUL_CODE == od_id),] 
  ocha_munic <- ocha_shape[which(ocha_shape$ADM2_PCODE == ocha_id), ]
  
  # Convert to st
  od_munic <- st_as_sf(od_munic)
  ocha_munic <- st_as_sf(ocha_munic)
  
  g <- ggplot() + #plot
  geom_sf(data = ocha_munic, color = "darkblue", fill = NA, linewidth = 1) +
  geom_sf(data = od_munic, color = "deeppink2", fill = NA, linewidth = 1)
  g <- g + theme_minimal()
  g <- g + theme(text = element_text(size = 14))
  g <- g + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
  g <- g + ggtitle(paste(ocha_munic$adm1_std,": ",ocha_munic$adm2_std))
  return(g)
}
```

```{r global_variables, include=FALSE}
PATH_SIRAJ <- "../../data/20240220/doi_10_5061_dryad_83nj1__v20190404/spatial_aggregate_non_timeseries/municip_aggregate_non_ts.csv" #Siraj et al. data sample
PATH_OD <- "../../data/20240612/Highest temporal resolution data_COLOMBIA_20021229_20191231.csv" #Open Dengue
PATH_OCHA_SHAPE <- "/Users/anna/Desktop/DD-Material/shapefiles/col-administrative-divisions-shapefiles/col_admbnda_adm2_mgn_20200416.shp" #shape file, OCHA
PATH_OD_SHAPE <- "/Users/anna/Desktop/DD-Material/shapefiles//OpenDengue_shapefile/Admin2(2011)_COL.shp" #shape file municipalities, Open Dengue
```

# Analysis Purpose

The purpose of this analysis is to link dengue incidence data from OpenDengue with spatiotemporal environmental data published in Siraj et al. (2018). Unfortunately, these two sources use two different identifier systems for municipalities. Siraj et al. (2018) uses identifiers provided by the Colombian government, while the OpenDengue database uses FAO GAUL identifiers. I was not able to find a resource that links these two identifier systems directly. However, I was able to retrieve shape files containing the FAU GAUL identifiers (corresponding author of the OpenDengue database, personal communication), and the Colombian identifier system (United Nations for the coordination of humanitarian affairs, OCHA).

A shape file is a simple, non-topological format for storing the geometric location and attribute information of geographic features. Here, the geographic features represent the boundaries of the municipalities in Colombia. I will us the geographic feature information as an aid to link the two identifier systems.


```{r read_data, include=FALSE}
env <- read.csv(PATH_SIRAJ, header = TRUE) #read Siraj sample data, containing processed spatiotemporal env. data
dengue <- read.csv(PATH_OD, header = TRUE) #read Dengue incidence data
dengue <- dengue[-which(dengue$T_res != "Week"),] #remove non-weekly resolution

ocha <- st_read(PATH_OCHA_SHAPE) #read OCHA shape file (2020)
od <- st_read(PATH_OD_SHAPE) #read OpenDengue shape file (2011)
```


# Check IDs

For the Siraj data, there is only one ID that is not present in the OCHA shape file - "Rea en Litigo" with an ID of 99999. However, there does not seem to be any dengue incidence data for "Rea en Litigo" anyways. For the Open Dengue data, all FAO GAUL IDs can be found in the provided shape file. This is not surprising, because the shape file was provided by the Open Dengue corresponding author.

For the Siraj data, I had to fall back on shape files that have been released after the publication of the data (the original files were not available). To see, whether there is consistency in municipality names, we will match the spatiotemporal environmental data from Siraj et al. (2018) with the OCHA data by ID, and compare the differences in naming with the help of a "string distance score": if two strings are identical, the score is 0. 

Overall, there is good agreement between the municipality names in Siraj et al. and the OCHA data (with the exception of the municipality with the ID 88564). Differences are mostly caused by additional information included in the municipality names in the Siraj at al data. Additionally, special characters are missing in the municipality names in the Siraj et al. data (confirmed by Amir Siraj, personal communication).

```{r ID_check}
ocha_id <- gsub("CO", "",ocha$ADM2_PCODE) #prepare OCHA ID
ocha_id <- as.integer(unique(ocha_id))
env_id <- as.integer(unique(env$ID_ESPACIA)) #prepare Siraj ID

table(env_id %in% ocha_id) #one Siraj ID not present in OCHA
absent_id <- setdiff(env_id, ocha_id)
env[which(env$ID_ESPACIA == absent_id),] # not present = REA EN LITIGIO 
table(dengue$adm_2_name[grep("REA", dengue$adm_2_name)]) #not present in incidence data

dengue_id <- as.integer(unique(dengue$FAO_GAUL_code)) #prepare OpenDengue ID
od_id <- as.integer(unique(od$GAUL_CODE))

table(dengue_id %in% od_id) #all OpenDengue IDs present in Open Dengue shape file
rm(absent_id, dengue_id, ocha_id, od_id, env_id)
```

```{r naming_check, include=FALSE}
#standardize Siraj et al. names
env_id <- subset(env, select = c("ID_ESPACIA", "NOM_MUNICI")) 
colnames(env_id) = c("ID", "env_name")
env_id$env_name_std <- tolower(stri_trans_general(env_id$env_name, "latin-ascii"))

#standardize OCHA names 
ocha_id <- data.frame("ID" = ocha$ADM2_PCODE, "ocha_name" = ocha$ADM2_ES)
ocha_id$ID <- as.integer(gsub("CO","",ocha_id$ID))
ocha_id$ocha_name_std <- tolower(stri_trans_general(ocha_id$ocha_name, "latin-ascii"))

#merge by ID
matched_ids <- merge(env_id, ocha_id, by = "ID")

#calculate distance in name strings 
matched_ids$score <- stringdist(matched_ids$env_name_std, matched_ids$ocha_name_std)
table(matched_ids$score) #majority only few mismatches 
matched_ids <- matched_ids[order(matched_ids$score, decreasing = TRUE),]
#print(matched_ids[which(matched_ids$score > 1),c("env_name_std", "ocha_name_std")])

ocha[which(ocha$ADM2_PCODE =="CO88564"),]
env[which(env$ID_ESPACIA == 88564),]

rm(env_id, ocha_id, matched_ids)
```

# Linking 
The linkage of spatiotemporal environmental data from Siraj et al. (2018) and dengue incidence data from the OpenDengue database comes with several challenges/caveats: 
- municipality names are not unique. Two municipalities (admin 2 level) can have the same name, but occur in two different departments (admin 1 level). 
- Siraj et al. (2018): special characters in the municipality names are missing 
- The FAO GAUL codes are not uniquely matched with individual municipalities: 2-4x assignments are presents (at a low frequency) too. The plots below demonstrate this problem. pink = region of one FAO GAUL code (OpenDengue shape file); blue = municipalities with same FAO GAUL Code (Ocha shape file). Thus, the shape files and FAO GAUL codes are only of limited use.  

```{r fao_gaul_code}
fao_gaul_code <- subset(dengue, select = c("FAO_GAUL_code", "full_name", "adm_2_name"))
fao_gaul_code <- fao_gaul_code[!duplicated(fao_gaul_code$full_name),]
print(nrow(fao_gaul_code)) #number municipalities 
print(length(unique(fao_gaul_code$FAO_GAUL_code))) #number FAO GAUL codes 

plot_gaul(gaul_id = 13521, od_shape = od, ocha_shape = ocha, dengue_data = dengue)
plot_gaul(gaul_id = 13441, od_shape = od, ocha_shape = ocha, dengue_data = dengue)
plot_gaul(gaul_id = 13516, od_shape = od, ocha_shape = ocha, dengue_data = dengue)

rm(fao_gaul_code)
```

## Approach

I match municipality and department names of the OpenDengue incidence data with the names available in the OCHA data. I will not use the Siraj et al. data directly, because it misses special characters. However, we have already established above that overall, the ID-assignment between the two data sets seems to be consistent. While the geographic feature information alone is not sufficient for matching (see examples above), I will use it as a quality control for incidences where the names do not match perfectly. 

```{r prep_link, echo=FALSE, warning=FALSE}
#generate data frame of municipalities with dengue incidences to be linked 
dengue_link <- subset(dengue, select = c("adm_1_name", 
                                         "adm_2_name", 
                                         "full_name",
                                         "FAO_GAUL_code"))

dengue_link <- dengue_link[!duplicated(dengue_link$full_name),]
dengue_link$ID <- c(1:nrow(dengue_link))

#standardize names 
dengue_link$adm2_std <- tolower(stri_trans_general(dengue_link$adm_2_name, "latin-ascii"))
dengue_link$adm1_std <- tolower(stri_trans_general(dengue_link$adm_1_name, "latin-ascii"))
ocha$adm2_std <- tolower(stri_trans_general(ocha$ADM2_ES, "latin-ascii"))
ocha$adm1_std <- tolower(stri_trans_general(ocha$ADM1_ES, "latin-ascii"))

link_df <- sapply(c(1:nrow(dengue_link)), function(k) #calculate matches 
  link_data(x = dengue_link[k,], xshape = od, y = ocha), simplify = F)

link_df <- do.call("rbind", link_df) #re=format matches 


```

### Step 1: Perfect Match

As a first step, we will match IDs, where both the standardized ( = lower case and latin ascii) municipality name and department name matches perfectly between the OCHA and the Siraj data (894 out of 1063 cases). 

```{r FilterCandidatesStep1}
#STEP 0: Generate variables for tracking ----
tobeLinked_ID <- unique(dengue_link$ID) #OpenDengue IDs to be linked
tobeLinked_df <- link_df #Data frame with OpenDeuge IDs ~ OCHA IDs link 

# STEP 1: Perfect Match ----
id <- which(link_df$adm1Score == 0 & link_df$adm2Score == 0)
length(unique(link_df$dengueID[id])) == nrow(link_df[id,]) #sanity check: all unique matches

final_link <- tobeLinked_df[id, ] 

#adjust remaining municipalities
tobeLinked_ID <- tobeLinked_ID[-which(tobeLinked_ID %in% final_link$dengueID)]
tobeLinked_df <- tobeLinked_df[-which(tobeLinked_df$dengueID %in% final_link$dengueID),]

table(tobeLinked_df$adm2Score, tobeLinked_df$adm1Score) #remaining Scores
```

### Step 2: Department MisMatch

Next, we will look at scenarios where the municipality name is a perfect match, but there are mismatches in the department names. 

The majority of these cases is caused by the departments norte de santander (OCHA)/norte santander (Open Dengue); valle del cauca (OCHA)/valle (Open Dengue); la guajira (OCHA)/guarija (Open Dengue). Visual inspection of the geographic feature information suggest that these cases match. Thus, we include them (981 out of 1063 matched). 

```{r FilterCandidatesStep2a}
#STEP 2: Perfect Munic. Score, Difference in Dep. ----
id <- which(tobeLinked_df$adm2Score == 0 & tobeLinked_df$adm1Score > 0)
table(tobeLinked_df$adm1_std[id])

#norte de santander (OCHA)  = norte santander (Open Dengue)
plot_gaul_byIDs(od_id = 14133, od_shape = od, ocha_id = "CO54206", ocha_shape = ocha)

#valle del cauca (OCHA) = valle (Open Dengue)
plot_gaul_byIDs(od_id = 13964, od_shape = od, ocha_id = "CO44090", ocha_shape = ocha)

#la guajira (OCHA) = guarija (Open Dengue )
plot_gaul_byIDs(od_id = 14378, od_shape = od, ocha_id = "CO76233", ocha_shape = ocha)


#STEP 2a: Dep. Santander, Valle, Guajira ----
id <- which(tobeLinked_df$adm2Score == 0 
            & tobeLinked_df$adm1_std %in% 
              c("norte de santander", "valle del cauca", "la guajira"))

#sanity check: only diff. in adm1 level
table(tobeLinked_df$adm2Score[id], tobeLinked_df$adm1Score[id]) 

#sanity check: constant difference in adm1 level 
table(tobeLinked_df$adm1Score[id], tobeLinked_df$adm1_std[id])

#sanity check: only unique matches 
length(unique(tobeLinked_df$dengueID[id])) == length(id)

final_link <- rbind(final_link, tobeLinked_df[id, ])

#adjust remaining municipalities
tobeLinked_df <- tobeLinked_df[-which(tobeLinked_df$dengueID %in% final_link$dengueID),]
tobeLinked_ID <- tobeLinked_ID[-which(tobeLinked_ID %in% final_link$dengueID)]

table(tobeLinked_df$adm2Score, tobeLinked_df$adm1Score) #remaining Scores
sum(tobeLinked_df$adm1Score >0 & tobeLinked_df$adm2Score == 0)
```

After accounting for the three departments, there are 14 remaining matches that have perfect municipality match, but mismatches in the department names. They are all rather dubious and I did not include them. 

```{r FilterCandidatesStep2b, include = FALSE}

#STEP 2b: Look at remaining adm2Score == 0 & adm1Score > 1----

for (i in tobeLinked_df$dengueID[which(tobeLinked_df$adm2Score == 0)]) {
  foo <- tobeLinked_df[which(tobeLinked_df$dengueID == i),]
  foo_id <- unique(foo$dengueID)
  dengue_id <- which(dengue_link$ID == foo_id)
  iter_res <- data.frame("dengueID" = rep(dengue_id, nrow(foo)),
                         "ochaID" = foo[["ADM2_PCODE"]],
                         "A2_OCHA" = foo[["adm2_std"]],
                         "A2_OD" = rep(dengue_link$adm2_std[dengue_id], nrow(foo)),
                         "A1_OCHA" = foo[["adm1_std"]],
                         "A1_OD" = rep(dengue_link$adm1_std[dengue_id], nrow(foo)))
  if(i == tobeLinked_df$dengueID[which(tobeLinked_df$adm2Score == 0)][1]) {
    res <- iter_res
  } else {
    res <- rbind(res, iter_res)
  }
}

print(res) #remaining links are dubious --> remove 

#adjust remaining municipalities
id <- which(tobeLinked_df$adm2Score == 0)
toremoveID <- tobeLinked_df$dengueID[id]
tobeLinked_df <- tobeLinked_df[-id,]
tobeLinked_ID <- tobeLinked_ID[-which(tobeLinked_ID %in% toremoveID)]

rm(toremoveID, foo, foo_id, iter_res, i, res, id, dengue_id) #clean up 
```
### Step 3: Municipality Mismatch 

Next, we will look at scenarios where the department is a perfect match, but there are mismatches in the municipality names (n = 41). Here, I also used visual inspection of the geographic features whether the match seems correct (i.e., partially similar municipality shape, considerable overlap) or whether it should be discarded (see plots below). 

```{r FilterCandidatesStep3, echo=FALSE}
#STEP 3: Perfect Department Match, MM in municipality---
sum(tobeLinked_df$adm1Score == 0 & tobeLinked_df$adm2Score > 0)

for (i in tobeLinked_df$dengueID[which(tobeLinked_df$adm1Score == 0)]) {
  foo <- tobeLinked_df[which(tobeLinked_df$dengueID == i),]
  foo_id <- unique(foo$dengueID)
  dengue_id <- which(dengue_link$ID == foo_id)
  iter_res <- data.frame("dengueID" = dengue_id,
                         "ochaID" = unique(foo[["ADM2_PCODE"]]),
                         "A2_OCHA" = unique(foo[["adm2_std"]]),
                         "A2_OD" = dengue_link$adm2_std[dengue_id],
                         "A1_OCHA" = unique(foo[["adm1_std"]]),
                         "A1_OD" = dengue_link$adm1_std[dengue_id])
  if(i == tobeLinked_df$dengueID[which(tobeLinked_df$adm1Score == 0)][1]) {
    res <- iter_res
  } else {
    res <- rbind(res, iter_res)
  }
}

#print(res)

badfitID <- c(2,3, 11, 13,14, 19:23,32, 35:43, 49)
goodfitID <- setdiff(c(1:nrow(res)), badfitID)

for (i in c(1:nrow(res))) { #visual inspection 
  id_dengue <- dengue_link$FAO_GAUL_code[which(dengue_link$ID == res$dengueID[i])]
  id_ocha <- res$ochaID[i]
  g <- plot_gaul_byIDs(od_id = id_dengue, 
                        ocha_id = id_ocha, 
                        od_shape = od, 
                        ocha_shape = ocha)
  if (i %in% badfitID)
    g <- g + ggtitle(paste(i, "bad"))
  else 
    g <- g + ggtitle(paste(i, "good"))
  print(g)
}

#add good fits to linked data 
final_link <- rbind(final_link, 
                    tobeLinked_df[which(tobeLinked_df$dengueID %in% res$dengueID[-badfitID]),])

#adjust remaining municipalities
tobeLinked_df <- tobeLinked_df[-which(tobeLinked_df$dengueID %in% res$dengueID),]
tobeLinked_ID <- tobeLinked_ID[-which(tobeLinked_ID %in% res$dengueID)]

table(tobeLinked_df$adm1Score, tobeLinked_df$adm2Score) #remaining scores 

rm(i, foo, foo_id, dengue_id, iter_res, id_dengue, id_ocha, badfitID, goodfitID, g, res)
```


```{r FilterCandidatesStep4, include=FALSE}
#STEP 4: look at remaining IDs manually

for (i in tobeLinked_df$dengueID) {
  foo <- tobeLinked_df[which(tobeLinked_df$dengueID == i),]
  foo_id <- unique(foo$dengueID)
  dengue_id <- which(dengue_link$ID == foo_id)
  foo <- subset(foo,select = c("adm1_std","adm2_std"))
  foo$geometry <- NULL
  colnames(foo) <- c("tbl_a1","tbl_a2")
  dengue_iter <- dengue_link[dengue_id, c("adm1_std","adm2_std")]
  
  print(foo)
  print(dengue_iter)
  print("######")
}
#no obivos good fit: linking finished 

rm(i, foo, foo_id, dengue_id, dengue_iter)
```

After these three steps, I matched 1009 out of 1063 municipalities (95 \%). The remaining matches were discarded. 

```{r StoreResults}
head(final_link)
nrow(final_link)/nrow(dengue_link) #able to link 95 % of municipalities. 

#subset to IDs 
final_link <- as.data.frame(subset(final_link, select = c("dengueID", "ADM2_PCODE")))
final_link$geometry <- NULL

#sanity check: any double entries 
any(duplicated(final_link$dengueID))
any(duplicated(final_link$ADM2_PCODE))

#store result
res <- merge(dengue_link, final_link, by.x = "ID", by.y = "dengueID")
write.table(res, file = "DD-AML-res-IDLinks.txt", 
            col.names = T, row.names = F, quote = F, sep = "\t")
```
